---
title: 'R package LakeEnsemblR: Basic Use and Sample Applications'
author:
   - "johannes.feldbauer@tu-dresden.de"
   - "tadhg.moore@dkit.ie"
   - "jorrit.mesman@unige.ch"
   - "ladwigjena@gmail.com"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
    toc_depth: 3
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{R package LakeEnsemblR: Basic Use and Sample Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- ALTERNATIVE FOR pdf
rmarkdown::pdf_document:
    includes:
      in_header: preamble-latex.tex
    number_sections: yes
    toc: yes
    toc_depth: 3
-->



```{r include=FALSE}
library(knitr)
library(kableExtra)
library(xtable)
library(LakeEnsemblR)
```

```{r include=FALSE}
# Store output format for later use
options(vignetteDocumentFormat = rmarkdown::all_output_formats("LakeEnsemblR_vignette.Rmd"))
options(xtable.comment = FALSE)
```

# Included models

LakeEnsemblR currently includes the following models: GLM (@hipsey_general_2019),
FLake (@mironov_flake_2008), GOTM (@umlauf_gotm_2005),
Simstrat (@goudsmit_application_2002), and MyLake (@saloranta_mylakemulti_year_2007).

# Introduction
LakeEnsemblR is an R package that lets you run multiple one-dimensional physical lake models.

The settings for a model run are controlled by one centralised, "master" configuration file in [YAML](https://yaml.org/) format. In this configuration file, you can set all the specifications for your model run, including start and end time, time steps, links to meteorological forcing and bathymetry files, etc. The package then converts these settings to the configuration files required by each model, through the `export_config` function. This sets up all models to run with the settings specified by the user, and the models are then run through the `run_ensemble` function. A netcdf file is created with the outputs of all the models, and the package provides functions to extract and plot this data.

Part of the design philosophy of LakeEnsemblR is that all input is given in a standardized format. This entails standard column names (which includes units), comma-separated ASCII files, and a DateTime format using the international standard format ([ISO 8601](https://en.wikipedia.org/wiki/ISO_8601)), which is `YYYY-mm-dd HH:MM:SS`, for example `2020-04-03 09:00:00`. In this document, we will explain what the required files are and in what format they need to be. We also advise you to look at the provided example files, and at the templates provided with the R package (to be found in `package/inst/extdata`, or extracted by the function `get_template`).

# Installation
The code of LakeEnsemblR is hosted on the AEMON-J Github page (https://github.com/aemon-j/LakeEnsemblR), and can be installed using the `devtools` package
```{}
devtools::install_github("aemon-j/LakeEnsemblR)
```

The package relies on multiple other packages that also need to be installed. Most notably, to run the multiple models, it requires the packages FLakeR, GLM3r, GOTMr, SimstratR, and MyLakeR. These packages run the individual models, and contain ways of running the models for the platforms Windows, MacOS, or UNIX, through either executables, or by having the model code in R. 

# The LakeEnsemblR configuration file
In this section, we go through the LakeEnsemblR configuration file. This file controls the settings with which the models are run. It is written in [YAML](https://yaml.org/) (Yaml Ain't Markup Language) format, and can be opened by text-editors such as Notepad or Notepad++. Although not needed to use LakeEnsemblR, you can read this file into R with the `configr` package (`read.config` function). Within LakeEnsemblR, we provide the `get_yaml_value` and `input_yaml` functions to get and input values into this file type.

There is an LakeEnsemblR configuration file provided in the example dataset in the package (`LakeEnsemblR::get_template("LakeEnsemblR_config")`) or you can download a copy from GitHub [here](https://github.com/aemon-j/LakeEnsemblR/blob/master/inst/extdata/feeagh/Feeagh_master_config.yaml).

## Location
The first section is "Location". Here you specify the name, latitude and longitude, elevation, maximum depth, and initial depth.

You also need to provide a link to the hypsograph (i.e. surface area per depth) file. As in the rest of the configuration file, all links to files are relative to the `folder` argument in the LakeEnsemblR functions (default is the R working directory). We strongly advise to set the working directory to the location of the LER config file, and link to the files relative to this directory (we explain this further in the chapter "Workflow" of this vignette). For example, if your hypsograph file is called `hypsograph.csv` and in the same folder as the LER config file, the corresponding line in the LER config file would look like
```{yaml}
  hypsograph: hypsograph.csv
```

The data needs to be a comma-separated file (.csv) where 0m is the surface and all depths are reported as positive, in meters. Area needs to be in meters squared. The column names *must* be `Depth_meter` and `Area_meterSquared`

Example of data:
```{csv}
Depth_meter,Area_meterSquared
0,3931000
1,3688025
2,3445050
3,3336093.492
4,3225992.455
5,3133491.11
6,3029720
...
```

## Time
In the "Time" section, you fill in the start and end date of the simulation, and the model time step. `time_step` indicates the model integration time step, so each time step that the model performs a calculation. 

## Config files
In this section, you link to the model-specific configuration files. Templates of these can be found in in the package (`get_template("FLake_config")`, `get_template("GLM_config")`, etc.) or you can download a copy from GitHub [here](https://github.com/aemon-j/LakeEnsemblR/blob/master/inst/extdata). The setup of LakeEnsemblR is such that you will usually not have to access these files, as most settings are regulated through the LER "master" config file. However, should you want to change some of the more specialised settings in each model, that is possible.

## Light
Here you can either give a value for Kw (light extinction coefficient) in 1/m, or give the link to a file, where you can vary Kw over time (template available). The file must be a comma-separated file (.csv) with the two columns `datetime` and 	`Extinction_Coefficient_perMeter` containing the date in ISO 8601 format (`YYYY-mm-dd HH:MM:SS`) and corresponding Kw value in 1/m.

## Observations
If you have observations of water temperature or ice cover, you can fill them in here. These will be used in plotting, and in case of water temperature potentially in initialising the temperature profile at the start of the simulation (see next section). 

For water temperature, the data needs to be a comma separated values (.csv) file where the datetime column is in the format `YYYY-mm-dd HH:MM:SS`. Depths are positive and relative to the water surface. Water temperature is in degrees Celsius. The column names *must* be `datetime`, `Depth_meter` and `Water_Temperature_celsius` (templates available).

Example of data:
```{}
datetime,Depth_meter,Water_Temperature_celsius
2004-01-05 00:00:00,0.9,6.97
2004-01-05 00:00:00,2.5,6.71
2004-01-05 00:00:00,5,6.73
2004-01-05 00:00:00,8,6.76
...
```

## Input
In the "Input" section, you give information about the initial temperature profile, meteorological forcing, and ice.

Firstly, you can give the initial temperature profile with which to start the simulation (link to .csv file, template available). If you leave it empty, the water temperature observations will be used, provided you have an observation on the starting time of your simulation.

Secondly, you give the link to the file with your meteorological data. The data needs to be a comma separated values (.csv) file where the datetime column is in the format `YYYY-mm-dd HH:MM:SS`. See Table 1 for the list of variables, units and column names. The `time_zone` setting has not been implemented yet.

```{r echo=FALSE, results='asis'}
# for html
x_html <- kable(met_input, caption = paste0("Table 1. Description of meteorological variables used ",
                                          "within LakeEnsemblR with units and required ",
                                          "column names."), row.names = FALSE, format = "html")
kable_styling(x_html, position = "left", font_size = 11)

# # for pdf output
# x_latex <- xtable(met_input, caption = paste0("Description of meteorological variables used ",
#                                           "within LakeEnsemblR with units and required ",
#                                           "column names."), format = "latex",
#                   align = c("p{0cm}", "p{1.8cm}", "p{0.5cm}", "p{7.2cm}", "p{5.1cm}"))
# print(x_latex, size = "\\scriptsize", include.rownames=FALSE )

```

Lastly, you can specify if you want to use the ice modules that are present in some of the models. 

## Inflows
Specify if you want to add inflows to your simulation. If yes, you need to link to a file with the inflow data. The data needs to be a comma separated values (.csv) file where the datetime column is in the format `YYYY-mm-dd HH:MM:SS`. Flow discharge, water temperature, and salinity need to be specified. The column names *must* be `datetime`, `Flow_metersCubedPerSecond`, `Water_Temperature_celsius`, and `Salinity_practicalSalinityUnits` (template available).

Example of data:
```{csv}
datetime,Flow_metersCubedPerSecond,Water_Temperature_celsius,Salinity_practicalSalinityUnits
2005-01-01 00:00:00,5.62,6.96,0.00
2005-01-02 00:00:00,2.32,6.00,0.00
2005-01-03 00:00:00,1.77,8.44,0.00
2005-01-04 00:00:00,4.64,7.27,0.00
...
```

## Output
In the "Output" section, you specify how the output should look like. Currently, only the netcdf option is in place. You can specify the depth interval of the output, and the frequency. Also specify what variables should be in the output (currently only "temp" and "ice_height").

## Biogeochemistry
Currently not implemented

## Model parameters
All models in LakeEnsemblR have different parameterizations. In this section, you can set the value of any parameter in one of the model-specific configuration files. 

You can give either only the name of the parameter, but if needed you can also provide the name of the section in which the parameter occurs, separated by a `/`. For example for GOTM's k_min parameter:
```{yaml}
  GOTM:
    k_min: 3.6E-6
```
or:
```{yaml}
  GOTM:
    turb_param/k_min: 3.6E-6
```

## Calibration
This section is used when calibrating the models. In the package, the run_LHC (Latin-Hypercube) and run_MCMC (Monte-Carlo Markov Chain) functions do this, but this is currently still in a beta-phase and this will be more clearly explained in a later version of this vignette.

# Workflow
In this chapter, we quickly show you how your workflow with LakeEnsemblR could look like. 

## Setting up a directory
First, you make an empty directory for the simulations of a specific lake. In this directory, you place the LakeEnsemblR configuration file, and the necessary LakeEnsemblR-format input files (e.g. meteorology, inflow, water temperature observations, hypsograph). Within this directory, you create empty folders for each model (FLake, GLM, GOTM, Simstrat, MyLake), and in here you place the model-specific configuration files, corresponding the information you put in the `config_files` section in the LER config file. 

## Export_config
Then, you run the `export_config` function, which exports the settings in the LER configuration file to the model-specific configuration files. It is possible to run parts of this function, such as `export_meteo`, if the meteorological forcing is the only thing you changed, but usually you will run `export_config`.

## Optional: changes in model-specific directories
Optionally, you can now go into the model-specific configuration files and make further changes. This should not be needed for regular simulations, especially since you can control parameters in the `model_parameters` section of the LER config file, but the option is there. For example, if you want to change the depth of the inflow in the Simstrat model, you could go to the Simstrat folders and make these changes manually (or through a script you wrote). It speaks for itself that these changes should be done only if you are sure this is what you need, and LakeEnsemblR does not provide further support for this.

## Run_ensemble
The next step would be to call `run_ensemble`, which runs all the models. A new folder called "output" is created, in which a netcdf-file will be put with all the results from the model runs. If you are interested, each model sub-folder will also have an "output" folder, with the model-specific model output.

## Post-processing
Now that the model simulations are finished, you can extract the data with `load_var`, or plot it with `plot_ensemble`. You can extract data from the netcdf file to do any further analysis. The `ncdf4` R package supports working with netcdf data in R, and the PyNcView software (https://sourceforge.net/projects/pyncview/) can be used to look at netcdf output.

We are currently working on a function to add new simulations (for example with a different meteorology file) to the same netcdf and including them in the analysis.

## Example model run
See below for an example run, and some post-processing of the data.
```{r eval = FALSE}
# Install packages - Ensure all packages are up to date - parallel development ongoing
#install.packages("devtools")
devtools::install_github("GLEON/GLM3r")
devtools::install_github(‘USGS-R/glmtools’, ref = ‘ggplot_overhaul’)
devtools::install_github("aemon-j/FLakeR", ref = “inflow”)
devtools::install_github("aemon-j/GOTMr")
devtools::install_github("aemon-j/gotmtools")
devtools::install_github("aemon-j/SimstratR")
devtools::install_github("aemon-j/LakeEnsemblR")
devtools::install_github("aemon-j/MyLakeR")

# Load libraries
library(gotmtools)
library(LakeEnsemblR)

# Set working directory with example data from Lough Feeagh, Ireland
template_folder <- system.file("extdata/feeagh", package= "LakeEnsemblR")
dir.create("example") # Create example folder
file.copy(from = template_folder, to = "example", recursive = TRUE)
setwd("example/feeagh") # Change working directory to example folder

# Set models & config file
model <- c("GLM",  "FLake", "GOTM", "Simstrat", "MyLake")
config_file <- "Feeagh_master_config.yaml"

# 1. Example - creates directories with all model setup
export_config(config_file = config_file, model = model, folder = ".")

# 2. Create meteo driver files
export_meteo(config_file = config_file, model = model)

# 3. Create initial conditions
start_date <- get_yaml_value(file = config_file, label =  "time", key = "start")

export_init_cond(config_file = config_file, 
                 model = model,
                 date = start_date,
                 print = TRUE)

# 4. Run ensemble lake models
wtemp_list <- run_ensemble(config_file = config_file,
                           model = c("FLake", "GLM", "GOTM", "Simstrat", "MyLake"),
                           return_list = TRUE)

```

Post-processing:

```{r eval = FALSE}

# Load libraries for post-processing
library(ggpubr)
library(ggplot2)

## Plot model output using gotmtools/ggplot2

# Extract names of all the variables in netCDF
ens_out <- "output/ensemble_output.nc"
vars <- gotmtools::list_vars(ens_out)
vars # Print variables

plist <- list() # Initialize empty list for storing plots of each variable
for(i in 1:5){
  p1 <- gotmtools::plot_vari(ncdf = ens_out,
                             var = vars[i],
                             incl_time = FALSE,
                             limits = c(0,25),
                             zlab = "degC")
  p1 <- p1 + scale_y_reverse() + #Reverse y-axis
    coord_cartesian(ylim = c(45,0))+ # ggplot2 v3.3 is sensitive to order of ylim
    ggtitle(vars[i]) + # Add title using variable name
    xlab("")+ # Remove x-label
    theme_bw(base_size = 18) # Increase font size of plots
  plist[[i]] <- p1
}

# Plot all model simulations
# install.packages("ggpubr")
g1 <- ggpubr::ggarrange(plotlist = plist, ncol = 1, common.legend = TRUE, legend = "right")
g1
ggsave("output/model_ensemble_watertemp.png", g1,  dpi = 300,width = 384,height = 300, units = "mm")

```

# Citation
See 
```{r eval=FALSE}
citation("LakeEnsemblR")
```
on how to cite this project. 

Although this information is included when running the function above, we would like to repeat that in case you use and cite LakeEnsemblR, you will also need to cite the individual models that you used in your simulations.

# References
