---
title: 'R package LakeEnsemblR: Basic Use and Sample Applications'
author:
   - "johannes.feldbauer@tu-dresden.de"
   - "tadhg.moore@dkit.ie"
   - "jorrit.mesman@unige.ch"
   - "ladwigjena@gmail.com"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{R package LakeEnsemblR: Basic Use and Sample Applications}
  %\VignetteEngine{knitr::rmarkdown}
---

<!-- ALTERNATIVE FOR html
rmarkdown::pdf_document:
    includes:
      in_header: preamble-latex.tex
    number_sections: yes
    toc: yes
    toc_depth: 3
-->


```{r include=FALSE}
# Store output format for later use
options(vignetteDocumentFormat = rmarkdown::all_output_formats("LakeEnsemblR_vignette.Rmd"))
```


```{r include=FALSE}
library(knitr)
library(xtable)
library(LakeEnsemblR)
```


# Setting up LakeEnsemblR

## Introduction
A key part of developing LakeEnsemblR was to develop a standardised format for model input data. This involved standard variable naming which includes units.

## DateTime formatting
This package uses international standard format for date and time (ISO 8601), which is `YYYY-mm-dd HH:MM:SS`. For example: `2020-04-03 09:00:00`.

## Time Zones
Currently this is not accounted for so the timezone used in input is the timezone of output data. It is on the list of things to do.

## Hypsograph data
The data needs to be a comma separated values (.csv) file where 0m is the surface and all depths are reported as positive. Area needs to be in meters squared. The column names *must* be `Depth_meter` and `Area_meterSquared`

Example of data:
```{}
Depth_meter,Area_meterSquared
0,3931000
1,3688025
2,3445050
3,3336093.492
4,3225992.455
5,3133491.11
6,3029720
...
```

## Temperature Profile data
The data needs to be a comma separated values (.csv) file where the datetime column is in the format `YYYY-mm-dd HH:MM:SS`. Depths are positive and relative to the water surface. Water temperature is in degrees Celsius. The column names *must* be `datetime`, `Depth_meter` and `Water_Temperature_celsius`

Example of data:
```{}
datetime,Depth_meter,Water_Temperature_celsius
2004-01-05 00:00:00,0.9,6.97
2004-01-05 00:00:00,2.5,6.71
2004-01-05 00:00:00,5,6.73
2004-01-05 00:00:00,8,6.76
...
```

## Meteorological data
The data needs to be a comma separated values (.csv) file where the datetime column is in the format `YYYY-mm-dd HH:MM:SS`. See table 1 for the list of variables, units and column names.

Table 1. Description of meteorological variables used within LakeEnsemblR with units and required column names.

| Description                      |  Units  | Column Name                                         | Status                                                                                                                                                            |
|----------------------------------|---------|-----------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| Downwelling longwave radaiation  | W/m2  | Longwave_Radiation_Downwelling_wattPerMeterSquared  | If not provided,it is calculated internally from air temperature, cloud cover and relative humidity/dewpoint temperature                                          |
| Downwelling shortwave radaiation | W/m2  | Shortwave_Radiation_Downwelling_wattPerMeterSquared | Required                                                                                                                                                          |
| Cloud cover                      | -     | Cloud_Cover_decimalFraction                         | If not provided,it is calculated internally from air temperature, short-wave radiation, latitude, longitude, elevation and relative humidity/dewpoint temperature |
| Air temperature                  | °C    | Air_Temperature_celsius                             | Required                                                                                                                                                          |
| Relative humidity                | %     | Relative_Humidity_percent                           | If not provided,it is calculated internally from air temperature and dewpoint temperature                                                                         |
| Dewpoint temperature             | °C    | Dewpoint_Temperature_celsius                        | If not provided,it is calculated internally from air temperature and relative humidity                                                                            |
| Wind speed at 10m                | m/s   | Ten_Meter_Elevation_Wind_Speed_meterPerSecond       | Either wind speed or u and v vectors is required                                                                                                                  |
| Wind direction at 10m            | °C    | Ten_Meter_Elevation_Wind_Direction_degree           | Not required, but if provided u and v vectors are calculated internally                                                                                                                                                      |
| Wind u-vector at 10m             | m/s   | Ten_Meter_Uwind_vector_meterPerSecond               | Either wind speed or u and v vectors is required                                                                                                                  |
| Wind v-vector at 10m             | m/s   | Ten_Meter_Vwind_vector_meterPerSecond               | Either wind speed or u and v vectors is required                                                                                                                  |
| Precipitation                    | m/s   | Precipitation_meterPerSecond                        | Not strictly required but is important for mass budgets in some models                                                                                                                                                      |
| Rainfall                         | m/s   | Rainfall_meterPerSecond                             | Required                                                                                                                                                          |
| Snowfall                         | m/day | Snowfall_meterPerDay                                | If not provided,it is calculated internally from rain when air temperature < 0 degC                                                                               |
| Sea level pressure               | Pa    | Sea_Level_Barometric_Pressure_pascal                | Not required                                                                                                                                                      |
| Surface level pressure           | Pa    | Surface_Level_Barometric_Pressure_pascal            | Required                                                                                                                                                          |
| Vapour pressure                  | mbar  | Vapor_Pressure_milliBar                             | If not provided,it is calculated internally from air temperature and relative humidity/dewpoint temperature                                                       |



## LakeEnsemblR YAML configuration file
There is an example yaml configuration provided in the example dataset in the package or you can download a copy from GitHub [here](https://github.com/aemon-j/LakeEnsemblR/blob/master/inst/extdata/feeagh/Feeagh_master_config.yaml).

You will need to update each of the required variables in the location block to reflect your own site.
```{}
location:
   name: Feeagh                                 # station name used in output [default=GOTM site]
   latitude: 53.9                               # latitude [degrees North; min=-90.0; max=90.0; default=0.0]
   longitude: -9.5                              # longitude [degrees East; min=-360.0; max=360.0; default=0.0]
   elevation: 15                                # elevation of lake surface above sea level [m]
   depth: 46.8                                  # maximum water depth [m; min=0.0; default=100.0]
   hypsograph: LakeEnsemblR_bathymetry_standard.csv                 # hypsograph [default=]
   init_depth: 46.8                             # initial height of lake surface relative to the bottom [m]
```

Then input the filepaths to the bathymetry, meteorlogical and water temperature profile observations file.
For first time users we would recommend to set up a folder with just these three files plus the LakeEnsemblR yaml configuration file.

Now you should be ready to run LakeEnsemblR on your site. 

# Running LakeEnsemblR

Once you have your hypsograph, water temperature observations and meteorological files prepared tunning LakeEnsemblR is relatively straightforward.

## Example model run
```{r eval = FALSE}
# Install packages - Ensure all packages are up to date - parallel development ongoing
#install.packages('devtools')
devtools::install_github('GLEON/GLM3r')
devtools::install_github(‘USGS-R/glmtools’, ref = ‘ggplot_overhaul’)
devtools::install_github('aemon-j/FLakeR', ref = “inflow”)
devtools::install_github('aemon-j/GOTMr')
devtools::install_github('aemon-j/gotmtools')
devtools::install_github('aemon-j/SimstratR')
devtools::install_github('aemon-j/LakeEnsemblR')
devtools::install_github('aemon-j/MyLakeR')

# Load libraries
library(gotmtools)
library(LakeEnsemblR)

# Set working directory
setwd('example') # Change working directory to example folder

# Set models & config file
model <- c('GLM',  'FLake', 'GOTM', 'Simstrat', 'MyLake')
config_file <- 'Feeagh_master_config.yaml'

# 1. Example - creates directories with all model setup
export_config(config_file = config_file, model = model, folder = '.')

# 2. Create meteo driver files
export_meteo(config_file = config_file, model = model)

# 3. Create initial conditions
start_date <- get_yaml_value(file = config_file, label =  "time", key = "start")

export_init_cond(config_file = config_file, 
                 model = model,
                 date = start_date,
                 print = TRUE)

# 4. Run ensemble lake models
wtemp_list <- run_ensemble(config_file = config_file,
                           model = c('FLake', 'GLM', 'GOTM', 'Simstrat', 'MyLake'),
                           return_list = TRUE)

```

## Post-processing
```{r eval = FALSE}

# Load libraries for post-processing
library(ggpubr)
library(ggplot2)

## Plot model output using gotmtools/ggplot2

# Extract names of all the variables in netCDF
ens_out <- 'output/ensemble_output.nc'
vars <- gotmtools::list_vars(ens_out)
vars # Print variables

plist <- list() # Initialize empty list for storing plots of each variable
for(i in 1:5){
  p1 <- gotmtools::plot_vari(ncdf = ens_out,
                             var = vars[i],
                             incl_time = FALSE,
                             limits = c(0,25),
                             zlab = 'degC')
  p1 <- p1 + scale_y_reverse() + #Reverse y-axis
    coord_cartesian(ylim = c(45,0))+ # ggplot2 v3.3 is sensitive to order of ylim
    ggtitle(vars[i]) + # Add title using variable name
    xlab('')+ # Remove x-label
    theme_bw(base_size = 18) # Increase font size of plots
  plist[[i]] <- p1
}

# Plot all model simulations
# install.packages('ggpubr')
g1 <- ggpubr::ggarrange(plotlist = plist, ncol = 1, common.legend = TRUE, legend = 'right')
g1
ggsave('output/model_ensemble_watertemp.png', g1,  dpi = 300,width = 384,height = 300, units = 'mm')

```

# References
